---
title: "Supplementary information for – The largest fully protected marine area in North America does not harm industrial fishing"
format: gfm
editor: visual
---

This repository stores code to support findings for a submitted manuscript titled "The largest fully protected marine area in North America does not harm industrial fishing". We provide here instruction to replicate the analysis over sample raw data and reproduce the results presented in the manuscript and in the supplementary materials.

This readme file is structured in two main chapters:

1.  Replicate the data: explains how data pre-processing occurs and raw data sources.

2.  Reproduce the results: explains how manuscript analysis, results, and figures are generated.

If reproducing the results using cleaned datasets is the main interest, the first section can be skipped entirely.

### dafishr package

The `dafishr` package was created to download, wrangle, and analyse Mexican VMS data. We strongly encourage to read `dafishr` documentation as well as this document. Many function are from the package and should be installed as:

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("CBMC-GCMP/dafishr")
```

## Replicate the data {#section-1}

Raw data from the Mexican Vessel Monitoring system are available through the "[Datos Abiertos](https://datos.gob.mx/busca/dataset/localizacion-y-monitoreo-satelital-de-embarcaciones-pesqueras)" initiative and can be downloaded and wrangled using the `dafishr` package. Here, we provide a quick intro.

### Step 1. Raw data download

First there are some packages that needs to be installed:

```{r, warning=FALSE, message=FALSE}
library(dafishr)
library(readr)
library(dplyr)
library(ggplot2)
library(future.apply)
library(purrr)
library(sf)
library(stringr)
library(CausalImpact)
```

Once installed the packages, we can use `dafishr` functions to download the last set of data.

```{r, eval=FALSE}
vms_download(2022)
```

The `vms_download` function will create a folder in the working directory called `VMS-data` that contains a series of `.csv` files from the Mexican SISMEP data from 2022. These data are raw information on geolocation of the industrial vessels. Different years can be selected at once using a vector inside the function. You can find more information on dafishr documentation `help(vms_download)`.

We can take a look at one of these files.

```{r, warning=FALSE, message=FALSE}
glimpse(read_csv('VMS-data/RLMSEP_2022/01. ENERO/01-15 ENE 2022.csv'))
```

The columns are:

-   `Nombre` = name of the vessel

-   `RNP` = unique vessel registration code

-   `Puerto base` = base port where vessels report catch

-   `Permisionario o Concesionario` = permit owner name or company name

-   `Fecha` = is the date of each geoposition

-   `Latitud` = is the WGS83 (4326) latitudinal degree

-   `Longitud` = is the WGS83 (4326) longitudinal degree

-   `Velocidad` = is the speed of the vessel at the time recorded

-   `Rumbo` = is the direction of navigation of the vessel in degrees

There are some evident parsing issues in speed and direction column, all these are considered and corrected in the pre-processing phase, as well as other errors.

### Step 2: Preprocessing

The pre-processing goes through a series of [steps](https://cbmc-gcmp.github.io/dafishr/articles/dafisr.html#downloading-raw-data-on-your-computer) that are wrapped inside a unique `preprocessing_vms` function. For simplicity, we use that wrapper here.

The function can be used on a single file, but we can loop it using `lapply` or even better use a parallel approach:

```{r, eval=FALSE}

## Create a list of files to process
files <- list.files("VMS-data/", recursive = T, pattern = ".csv", full.names = T)

## Set up a parallel session 
plan(multisession, workers = 2) ## Set Cores according to laptop characteristics

## Use the preprocessing_vms function wrapper
future_lapply(files, preprocessing_vms, future.seed = NULL)

```

The `future_lapply` function will loop the `preprocessing_vms` function on the list of files that were downloaded in the `VMS-data` folder and save cleaned files in a `.fst` format (see specs [here](https://www.fstpackage.org/)) in a preprocessed folder that is automatically created in the working directory.

If you apply the preprocessing on the full scale of the VMS data (from 2008 to 2022) it will take a long time to process on a personal computer. However, the approach of keeping small chunks of data separated allows to complete all the analysis without the need of a significant computing power.

We can load a file to check the results:

```{r, warning=FALSE, message=FALSE}
## Load a file to check 
vms <- fst::read_fst("preprocessed/vms_2022_1_1_15_preprocessed.fst")

glimpse(vms)

```

Column names are now in English, the parsing is now as expected and several other columns are added:

-   `file_name` = the name of the raw file which is useful for future references and bug/errors detection

-   `location` = has two levels "port_visit", "at_sea", meaning that the vessels was at port or navigating at sea respectively. Information on how this is obtained can be found [here](https://cbmc-gcmp.github.io/dafishr/reference/join_ports_locations.html).

-   `zone` = can be "open area" meaning that the vessel was outside an MPA polygon, or it can be a name of an MPA where the vessels was operating. The intersection method with all the MPA in Mexico can be found [here](https://cbmc-gcmp.github.io/dafishr/reference/join_mpa_data.html).

-   `mpa_decree` = the decree of the MPA is coded as "NA" if absent, "PN" for National Park, "PMN" for National Marine Park, "RB" Biosphere reserve, "APFF" Area of Protection for the Flora and Fauna.

-   `state` = the state that the MPA belongs to administratively

-   `municipality` = the municipality that the MPA belongs to administratively

-   `region` = the region that the MPA belongs to administratively

We use this intersection to discriminate vessels that historically were fishing in Revillagigedo and vessels that did not.

### Step 3: Modeling VMS data

Now that the preprocessing is over, we can model VMS data according to the speed to have a sense of where vessels were probably fishing. Modeling details can be found [here](https://cbmc-gcmp.github.io/dafishr/reference/model_vms.html) and the full code [here](https://github.com/CBMC-GCMP/dafishr/blob/HEAD/R/model_vms.R).

Here, we first create a list of all the preprocessed files and then we loop using `map_dfr` function from the `purrr` package the `model_vms` function. We add an additional parsing rule to the `RNP` column as when reading back some of the files it is sometimes parsed as character drawing an error in the final merge.

The approach can also be parallelized using `furrr` version of the function: `future_map_dfr`.

```{r, warning=FALSE, message=FALSE, eval=FALSE}
## Create a list of all the files that were preprocessed

files_preprocessed <- list.files("preprocessed/", recursive = T, pattern = ".fst", full.names = T)

## Model files and create a new data frame
vms_modeled  <- map_dfr(files_preprocessed, 
                        function(x)
                              fst::read_fst(x) |>
                              mutate(RNP = as.character(RNP)) |> 
                              model_vms())


```

We can now create a plot of the results to see where the vessels were active.

```{r, warning=FALSE, message=FALSE, eval=FALSE}
## Plot to observe
vms_modeled |> 
      filter(vessel_state == "hauling") |> 
      st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |> 
      ggplot() +
      geom_sf(pch = ".") +
      geom_sf(data = dafishr::all_mpas, fill = NA, col = "red") +
      coord_sf(xlim = c(min(vms_modeled$longitude), max(vms_modeled$longitude)), 
               ylim = c(min(vms_modeled$latitude), max(vms_modeled$latitude)))
```

## Reproduce the results

The road from raw data to processed files of all the historical data is long and we are available for questions and details on how to fully reproduce all the details. Beware, however, that all significant steps are presented above and now we give all relevant information on data and analysis to properly assess caveats and limitations of the study.

### Fishing activity inside Revillagigedo MPA Polygon

Using monthly data that were modeled using the `dafishr` package as described above, we can now discriminate potential fishing activity inside MPA polygons and see how it changed over time.

First however, we want to filter only the vessels that had a permit to fish Tuna, Sharks, and Marlins using longlines or purse seines. This results in a pelagic fleet, the one that might be mostly affected by the establishment of Revillagigedo MPA.

We load the permit vessel list from the `dafishr` package, more description of the dataset is available [here](https://cbmc-gcmp.github.io/dafishr/reference/pelagiv_vessels_permits.html).

```{r, warning=FALSE, message=FALSE}
permits <- dafishr::pelagic_vessels_permits |> 
      mutate(str_split(vessel_name, "\\(")) |> # some names adjustments 
      pull(vessel_name) |> 
      unique()
```

#### Fishing activity dataset

In this repo we make available the results of the preprocessing and modeling of VMS dataset by the `dafishr` package.

```{r, warning=FALSE, message=FALSE}
glimpse(read_rds("outputs/month_vessel_hrs.RDS"))
```

The dataset is structured by year, month, name of the vessel (`vessel_name`), the zone of activity and the duration of the potential fishing activity in hours (`hrs`).

From this dataset we can create: 1) a `vessels_list` of all the vessels found active in our study area; and 2) a `revi_list` which is all the vessels that were found active in the Revillagigedo area. The latter was obtained by intersecting the VMS data modeled with the Mexican MPAs polygons (available in `dafishr::all_mpas`)

Then, we create a vessels list and a list of all the vessels that were found to fish in Revillagigedo area.

```{r, warning=FALSE, message=FALSE}
vessels_list <- read_rds("outputs/month_vessel_hrs.RDS") %>% 
      pull(vessel_name) %>% 
      unique()

revi_list <- read_rds("outputs/month_vessel_hrs.RDS") %>% 
      filter(str_detect(zone, "Revillagigedo")) %>% 
      pull(vessel_name) %>% 
      unique()

```

We now create a `revispatial` object, by filtering the Revillagigedo zone, wrangling dates and calculating the effort by dividing the hours of activity by the number of vessels active.

```{r, warning=FALSE, message=FALSE}
revispatial <- readRDS("outputs/month_vessel_hrs.RDS") %>%  
      filter(str_detect(zone, "Revillagigedo")) %>%  
      group_by(year, month, vessel_name) %>% 
      summarise(hrs = sum(hrs)) %>% 
      group_by(year, month) %>%  
      summarise(vessel = n_distinct(vessel_name), hrs = sum(hrs/24)) %>% 
      mutate(effort = hrs/vessel, 
             date = as.Date(paste0(year, "-", month, "-01"), "%Y-%m-%d"))

```

Finally, we can plot the results. This plot reproduces Figure 1A in the manuscript.

```{r, warning=FALSE, message=FALSE}

revispatial %>%
      mutate(period = ifelse(date < "2017-11-01", "Before", "After")) |>
      ggplot(aes(x = date, y = effort)) +
      geom_point(
            pch = 21,
            fill = "gray80",
            col = "black",
            alpha = .4
      ) +
      labs(x = "",
           y = "Average Fishing Effort",
           title = "Fishing activity within MPA polygon") +
      geom_hline(yintercept = 0) +
      geom_vline(xintercept = as.Date("2017-11-01"),
                 col = "red",
                 size = 1) +
      geom_line(aes(group = period)) +
      geom_smooth(
            aes(group = period),
            method = "gam",
            method.args = list(family = Gamma(link = "log")),
            se = F
      ) +
      scale_color_manual(values = c("#FFC107", "#1E88E5")) +
      scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
      theme(
            legend.position = "",
            legend.text = element_text(face = "bold"),
            panel.background = element_rect(fill = NA),
            plot.background = element_rect(fill = NA),
            panel.grid.major = element_line(color = "gray95"),
            text = element_text(color = "gray50"),
            axis.text = element_text(color = "gray50"),
            axis.title = element_text(color = "gray50"),
            axis.text.x = element_text(angle = 90),
            strip.background = element_blank(),
            axis.line.y = element_line(color = "black")
      )



```
